Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-20 ind_files/99_Appendix_D_examples.tex
Correct minor typo

In frequency space, the complete mesoscopic description for the stochastic dynamics can be similarly calculated in terms of the matrices $\mathbf{M}$ and $\mathbf{T}$, and will yield:
\begin{equation}
dp_i =  N_K(t)\left( \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i - \frac{1}{KN_K}\left[(\mathbf{Tp})_i - \mathbf{p}\cdot\mathbf{Tp}\right]p_i\right)dt + \frac{1}{\sqrt{K}}\left(p_i(\mathbf{Tp})_i - p_i^2(\mathbf{Tp})_i\right)^{1/2}dW_t 
\end{equation}

@@

For this model, we can in fact quantitatively derive the effects of noise-induced selection by explicitly calculating each term of equation \eqref{nD_stochastic_RM}. Let $p = x_1/(x_1 + x_2)$ be the frequency of type 1 individuals and let $q = 1-p$. Then, it is easy to check by direct substitution of our functional forms that we have
\begin{align*}
	\overline{w} &= w_1(\mathbf{x}) = w_2(\mathbf{x})\\
	\overline{\tau} &= 1 + x_1 + x_2(q-p)\\
	\mu Q_i(\mathbf{p}) &= \mu p_{j}\textrm{ , where $i \neq j$}\\
	\mu(Q_1(\mathbf{p}) - p\left(\sum\limits_{j=1}^{2} Q_j(\mathbf{p})\right)) &= \mu(q-p)
\end{align*}
And thus, equation \eqref{nD_stochastic_RM} becomes
\begin{equation}
	\label{App_stoch_LV_RM_eqn}
	dp =  \left[\frac{2}{K}p^2q + \mu\left(1-\frac{1}{KN_K}\right)\left(q-p\right)\right]dt + q\sqrt{A^{+}_1}dW_t^{(1)} - p\sqrt{A^{+}_2}dW_t^{(2)}
\end{equation}
where $A_i^{+} = x_i\tau_i(\mathbf{x}) +\mu Q_i(\mathbf{p})$ and each $W^{(i)}_t$ is an independent Wiener process. This equation clearly shows the biasing effect of noise-induced selection in the first component of the $dt$ term of the RHS. Since $p^2(1-p) > 0$ for $p \in (0,1)$, this term always tends to increase the fraction of type 1 individuals in the population. Note that the difference in fitness between the two types remains zero if every competition term (\emph{i.e.} every $n_in_j/K$ term) is multiplied by some constant $\alpha > 0$ parameterizing the strength of competition, meaning that the two types still have equal fitness. However, this constant affects the strength of noise-induced selection, and the corresponding term in equation \ref{App_stoch_LV_RM_eqn} becomes $2\alpha p^2q/K$ instead of $2p^2q/K$. Thus, for $\alpha > 1$, the strength of noise-induced selection (and thus the extent to which the distribution of types in the population is biased in favor of type 1 in plots like \ref{App_stoch_LV_dens_plots}) can be made arbitrarily high simply by modulating the strength of competition. The second component in the $dt$ term captures the effects of mutations, and simply reflects the fact that we assumed that $\textrm{(type }1) \to \textrm{(type }2)$ and $\textrm{(type }2) \to \textrm{(type }1)$ mutations occur at the same rate $\mu$, and thus, the net effect of mutational effects depends on the difference between the frequencies of the two types and flows towards the type with lower frequency. Finally, the two $dW^{(i)}_t$ terms are non-directional and vanish upon taking an expectation over the probability space, and therefore have no net contribution other than `blurring out the results' if we look at the dynamics averaged over many realizations.

>>>>

In frequency space, the complete mesoscopic description for the stochastic dynamics can be similarly calculated in terms of the matrices $\mathbf{M}$ and $\mathbf{T}$, and will yield:
\begin{equation}
\resizebox{\textwidth}{!}{$\displaystyle dp_i =  N_K(t)\left( \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i - \frac{1}{KN_K}\left[(\mathbf{Tp})_i - \mathbf{p}\cdot\mathbf{Tp}\right]p_i\right)dt + \frac{1}{\sqrt{K}}\left(p_i(\mathbf{Tp})_i - p_i^2\sum\limits_{j=1}^{m}(\mathbf{Tp})_j\right)^{1/2}dW_t$} 
\end{equation}

@@

For this model, we can in fact quantitatively derive the effects of noise-induced selection by explicitly calculating each term of equation \eqref{nD_stochastic_RM}. Let $p = x_1/(x_1 + x_2)$ be the frequency of type 1 individuals and let $q = 1-p$. Then, it is easy to check by direct substitution of our functional forms that we have
\begin{align*}
	\overline{w} &= w_1(\mathbf{x}) = w_2(\mathbf{x})\\
	\overline{\tau} &= 1 + x_1 + x_2(q-p)\\
	\mu Q_i(\mathbf{p}) &= \mu p_{j}\textrm{ , where $i \neq j$}\\
	\mu(Q_1(\mathbf{p}) - p\left(\sum\limits_{j=1}^{2} Q_j(\mathbf{p})\right)) &= \mu(q-p)
\end{align*}
And thus, equation \eqref{nD_stochastic_RM} becomes
\begin{equation}
	\label{App_stoch_LV_RM_eqn}
	dp =  \left[\frac{2}{K}p^2q + \mu\left(1-\frac{1}{KN_K}\right)\left(q-p\right)\right]dt + \frac{1}{\sqrt{K}N_K(t)}\left[q\sqrt{A^{+}_1}dW_t^{(1)} - p\sqrt{A^{+}_2}dW_t^{(2)}\right]
\end{equation}
where $A_i^{+} = x_i\tau_i(\mathbf{x}) +\mu Q_i(\mathbf{p})$ and each $W^{(i)}_t$ is an independent Wiener process. This equation clearly shows the biasing effect of noise-induced selection in the first component of the $dt$ term of the RHS. Since $p^2(1-p) > 0$ for $p \in (0,1)$, this term always tends to increase the fraction of type 1 individuals in the population. Note that the difference in fitness between the two types remains zero if every competition term (\emph{i.e.} every $n_in_j/K$ term) is multiplied by some constant $\alpha > 0$ parameterizing the strength of competition, meaning that the two types still have equal fitness. However, this constant affects the strength of noise-induced selection, and the corresponding term in equation \ref{App_stoch_LV_RM_eqn} becomes $2\alpha p^2q/K$ instead of $2p^2q/K$. Thus, for $\alpha > 1$, the strength of noise-induced selection (and thus the extent to which the distribution of types in the population is biased in favor of type 1 in plots like \ref{App_stoch_LV_dens_plots}) can be made arbitrarily high simply by modulating the strength of competition. The second component in the $dt$ term captures the effects of mutations, and simply reflects the fact that we assumed that $\textrm{(type }1) \to \textrm{(type }2)$ and $\textrm{(type }2) \to \textrm{(type }1)$ mutations occur at the same rate $\mu$, and thus, the net effect of mutational effects depends on the difference between the frequencies of the two types and flows towards the type with lower frequency. Finally, the two $dW^{(i)}_t$ terms are non-directional and vanish upon taking an expectation over the probability space, and therefore have no net contribution other than `blurring out the results' if we look at the dynamics averaged over many realizations.
